Crystal as a thermodynamic computer

### 2024-09-05 paper/main.tex
fix notation issues

\newcommand{\jinguo}[1]{[{\color{blue}{JGL: #1}}]}
\newcommand{\ym}[1]{[{\color{red}{YM: #1}}]}
\newcommand{\s}{\mathbf {s}}
\newcommand{\net}[1]{{\textsc{#1}}}
\newcommand{\argmin}{\mathop{\mathrm{argmin}}\limits}
\newcommand{\Eq}[1]{Eq.~(\ref{#1})}
\newcommand{\Fig}[1]{Fig.~\ref{#1}}

@@

Improving the energy efficiency of computation is a major challenge in the development of future computing systems.
The thermodynamic computer (TC)~\cite{conte2019thermodynamic} is a new type of computer that uses thermodynamics to perform computation.
Instead of performing the computation deterministically, TCs are stochastic and are driven by thermodynamics.
Physical implementations of TCs include DNA machines~\cite{tanaka2005design} and Ising machines~\cite{gu2012encoding, lucas2014ising, mohseni2022ising, fabre2014optical}.
% memristors~\cite{yang2013memristive, wang2017memristors, kumar2017chaotic, wang2018fully},

@@

\begin{equation}
H = \sum_{i < j} J_{ij}s_us_v + \sum_{i=1}^{n}h_i s_i
\end{equation}

@@

Previous studies have shown that the Ising model is universal for computation~\cite{Aadit2022, onizawa2020design, camsari2017stochastic}, meaning that any logic circuit can be encoded into the Ising model and solved by finding the ground state of the Ising model. A universal set logic gate gadgets are designed, by concatenating which any logic circuit can be constructed.
For example the logic expression $z = \neg (a \land b)$ can be encoded into the Ising model by the following Hamiltonian:
\begin{equation}
    \begin{split}
        H_{c = a \land b} &= s_a s_b -2 s_a s_c -2 s_bs_c + s_a + s_b -2 s_c\\
        H_{z = \neg c} &= s_c s_z\\
        H_{z = \neg (a \land b)} &= H_{z = \neg c} + H_{c = a \land b}
    \end{split}
\end{equation}
where, the variable $c$ is an ancilla spin.
A CA can be viewed as a translational invariant logic circuit, and thus can be encoded into the Ising model. The ground states of the Ising model encode the correspondence between the inputs and the outputs of the CA.
Directly synthesizing the logic circuit for the universal $110$ rule (\Cref{eq:110rule}) requires at least 3 ancilla spins, which is far from optimal. By converting the gadget finding problem to a linear programming problem~\cite{Kolman1995}, we propose a provable optimal method to encode any CA rule into the Ising model, which is described in \Cref{sec:gadget-design}.
The proposed gadget for the $110$ rule is shown in \Cref{Surface-programmable-thermodynamic computer}, which only requires 1 ancilla spin.
The interaction energy and onsite energy of the rule 110 gadget are given below, where $p, q, r$ and inputs and $q'$ is the output in \Cref{Surface-programmable-thermodynamic computer}, respectively, while spin $A$ is the auxiliary spin.
\begin{equation}
J = \begin{blockarray}{cccccc}
& p & q & r & q^\prime & A\\
\begin{block}{c(ccccc)}
p & ~\cdot~ & ~1~ & ~1~ & ~2~ & ~3~\\
q & \cdot & \cdot & 2 & 2 & 5\\
r & \cdot & \cdot & \cdot & 2 & 5\\
q^\prime & \cdot & \cdot & \cdot & \cdot & 6\\
A & \cdot & \cdot & \cdot & \cdot & \cdot\\
\end{block}
\end{blockarray}
~~, ~~h = \begin{blockarray}{cc}
& \phantom{x} \\
\begin{block}{c(c)}
p & ~1~\\
q & 2\\
r & 2\\
q^\prime & 2\\
A & 5\\
\end{block}
\end{blockarray}
\end{equation}
%Since a copy gate\jinguo{do we really use copy gates?} \ym{No need in practice, but indeed used in Fig.1a} can be simply described as two spins with an interaction $J_{u, v}=1$, and by applying the method of connecting logic gates~\cite{Aadit2022}, we can concatenate the rule 110 gadget with some copy gates, ultimately forming
By concatenating the rule 110 gadgets, we can construct a 2-dimensional TC that implements the rule 110 CA, as shown in \Cref{Surface-programmable-thermodynamic computer}.
The gadget we constructed based on the Rule 110 CA naturally possesses Turing completeness. Therefore, this 2-dimensional TC is capable of universal computing.

@@

With the proposed Ising model, we are interested in how efficiently we can drive the system to the ground state, which encodes the output of the CA.
Since the natural cooling process is highly nontrivial to model, we consider a theoretical model for cooling, which is simulated annealing (SA)~\cite{Kirkpatrick1983}.
\jinguo{random fluctuations may be linear in time considering the parallelism of the system.}
It is known that when the system and heat bath are at thermal equilibrium, 
the probability of a spin configuration $\mathbf{s}$ is given by the Boltzmann distribution:

\jinguo{please do not use the ket bra notation}
\begin{equation}
    \mathbf{s} = \frac{1}{Z}\exp\left(- \frac{E_{\mathbf{s}}}{k_bT}\right)
\end{equation}

where $Z = \sum_{\mathbf{s}} \exp\left(-\frac{E_{\mathbf{s}}}{k_bT}\right)$ is the partition function, 
$E_{\mathbf{s}}$ is the energy of the each of $2^n$ spin configuration, labeled by $\mathbf{s} = (s_1, s_2, \cdots, s_n)$, $k_b$ is the Boltzmann constant, and $T$ is the temperature of the heat bath. 
In the following, we set $k_b = 1$.

We employ the Metropolis-Hastings algorithm~\cite{metropolis1953equation} to characterize the thermodynamic process of the TC system as modulated by the heat bath.
This algorithm perform a random walk in the state space of the TC, each step we try to flip a spin,
and the probability of accepting a move from state $\mathbf{s}$ to $\mathbf{s}^\prime$ is given by:

\begin{equation}
    P_{\text{accept}} = \frac{1}{1+\exp\left(\frac{E_{\mathbf{s}^\prime} - E_{\mathbf{s}}}{T}\right)}
\end{equation}

This accepting probability follows from the glauber dynamics~\cite{glauber1963time}.
It can be shown that this process is markovian and ergodic, and the system will eventually reach the thermal equilibrium.
To determine the speed of the convergence, we should notice the spectral gap of the transition matrix of the Markov chain, which is defined as

@@

Denote $S$ to be the length of the material along the ``Space'' axis, $D$
to be the length of the material along the ``Time'' axis. 
Let us define a simpler but equivalent energy model on a 2D lattice of size $S \times D$, whose hamiltonian
is given by
\begin{equation}\label{eq:toy-hamiltonian}
H = \sum_{t=1}^{D-1}\sum_{i=1}^{S} (n_{i+1, t} \land \neg(n_{i, t} \land \neg n_{i-1, t})) \veebar n_{i,t+1},
\end{equation}

@@

\section{The implmentation in Rydberg atoms array}\label{physical-model-rydberg-atoms-array}
Let us consider a Rydberg atoms array with a two dimensional embedding. The coordinate of the atom $v$ is denoted as $\mathbf{r}_v$.
The Rydberg Hamiltonian~\cite{Nguyen2023} is defined as

>>>>

\newcommand{\jinguo}[1]{[{\color{blue}{JGL: #1}}]}
\newcommand{\ym}[1]{[{\color{red}{YM: #1}}]}
\newcommand{\s}{\mathbf {s}}
\newcommand{\net}[1]{{\textsc{#1}}}
\newcommand{\vs}{{\mathbf{s}}}
\newcommand{\vn}{{\mathbf{n}}}
\newcommand{\argmin}{\mathop{\mathrm{argmin}}\limits}
\newcommand{\Eq}[1]{Eq.~(\ref{#1})}
\newcommand{\Fig}[1]{Fig.~\ref{#1}}

@@

Improving the energy efficiency of computation is a major challenge in the development of future computing systems.
The thermodynamic computer (TC)~\cite{conte2019thermodynamic, hylton2021vision} is a new type of computer that uses thermodynamics to perform computation.
Instead of performing the computation deterministically, TCs are stochastic and are driven by thermodynamics.
Physical implementations of TCs include DNA machines~\cite{tanaka2005design} and Ising machines~\cite{gu2012encoding, lucas2014ising, mohseni2022ising, fabre2014optical}.
% memristors~\cite{yang2013memristive, wang2017memristors, kumar2017chaotic, wang2018fully},

@@

\begin{equation}
E(\vs) = \sum_{i < j} J_{ij}s_us_v + \sum_{i=1}^{n}h_i s_i
\end{equation}

@@

Previous studies have shown that the Ising model is universal for computation~\cite{Aadit2022, onizawa2020design, camsari2017stochastic}, meaning that any logic circuit can be encoded into the Ising model and solved by finding the ground state of the Ising model. A universal set logic gate gadgets are designed, by concatenating which any logic circuit can be constructed.
For example the logic expression $z = \neg (a \land b)$ can be encoded into the Ising model by the following Hamiltonian:
\begin{equation}
    \begin{split}
        E_{c = a \land b} &= s_a s_b -2 s_a s_c -2 s_bs_c + s_a + s_b -2 s_c\\
        E_{z = \neg c} &= s_c s_z\\
        E_{z = \neg (a \land b)} &= E_{z = \neg c} + E_{c = a \land b}
    \end{split}
\end{equation}
where, the variable $c$ is an ancilla spin.
A CA can be viewed as a translational invariant logic circuit, and thus can be encoded into the Ising model. The ground states of the Ising model encode the correspondence between the inputs and the outputs of the CA.
Directly synthesizing the logic circuit for the universal $110$ rule (\Cref{eq:110rule}) requires at least 3 ancilla spins, which is far from optimal. By converting the gadget finding problem to a linear programming problem~\cite{Kolman1995}, we propose a provable optimal method to encode any CA rule into the Ising model, which is described in \Cref{sec:gadget-design}.
The proposed gadget for the $110$ rule is shown in \Cref{Surface-programmable-thermodynamic computer}, which only requires 1 ancilla spin.
The interaction energy and onsite energy of the rule 110 gadget are given below, where $p, q, r$ and inputs and $q'$ is the output in \Cref{Surface-programmable-thermodynamic computer}, respectively, while spin $A$ is the auxiliary spin.
\begin{equation}
J = \begin{blockarray}{cccccc}
& p & q & r & q^\prime & A\\
\begin{block}{c(ccccc)}
p & ~\cdot~ & ~1~ & ~1~ & ~2~ & ~3~\\
q & \cdot & \cdot & 2 & 2 & 5\\
r & \cdot & \cdot & \cdot & 2 & 5\\
q^\prime & \cdot & \cdot & \cdot & \cdot & 6\\
A & \cdot & \cdot & \cdot & \cdot & \cdot\\
\end{block}
\end{blockarray}
~~, ~~h = \begin{blockarray}{cc}
& \phantom{x} \\
\begin{block}{c(c)}
p & ~1~\\
q & 2\\
r & 2\\
q^\prime & 2\\
A & 5\\
\end{block}
\end{blockarray}
\end{equation}
%Since a copy gate\jinguo{do we really use copy gates?} \ym{No need in practice, but indeed used in Fig.1a} can be simply described as two spins with an interaction $J_{u, v}=1$, and by applying the method of connecting logic gates~\cite{Aadit2022}, we can concatenate the rule 110 gadget with some copy gates, ultimately forming
By concatenating the rule 110 gadgets, we can construct a 2-dimensional TC that implements the rule 110 CA, as shown in \Cref{Surface-programmable-thermodynamic computer}.
The gadget we constructed based on the Rule 110 CA naturally possesses Turing completeness. Therefore, this 2-dimensional TC is capable of universal computing.

@@

With the proposed Ising model, we are interested in how efficiently we can drive the system to the ground state, which encodes the output of the CA.
Since the natural cooling process is highly nontrivial to model, we consider a theoretical model for cooling, which is simulated annealing (SA)~\cite{Kirkpatrick1983}.
\jinguo{random fluctuations may be linear in time considering the parallelism of the system.}
It is known that when the system and heat bath are at thermal equilibrium, 
the probability of a spin configuration $\mathbf{s}$ is given by the Boltzmann distribution:
\begin{equation}
    p(\mathbf{s}) = \frac{1}{Z}\exp\left(- \frac{E({\mathbf{s}})}{k_bT}\right),
\end{equation}
where $Z = \sum_{\mathbf{s}} \exp\left(-\frac{E({\mathbf{s}})}{k_bT}\right)$ is the partition function, 
$E({\mathbf{s}})$ is the energy of the each of $2^n$ spin configuration, labeled by $\mathbf{s} = (s_1, s_2, \cdots, s_n)$, $k_b$ is the Boltzmann constant, and $T$ is the temperature of the heat bath. 
In the following, we set $k_b = 1$.

We employ the Metropolis-Hastings algorithm~\cite{metropolis1953equation} to characterize the thermodynamic process of the TC system as modulated by the heat bath.
This algorithm perform a random walk in the state space of the TC, each step we try to flip a spin,
and the probability of accepting a move from state $\mathbf{s}$ to $\mathbf{s}^\prime$ is given by:
\begin{equation}
    P_{\text{accept}} = \frac{1}{1+\exp\left(\frac{E({\mathbf{s}^\prime}) - E({\mathbf{s}})}{T}\right)}.
\end{equation}
This accepting probability follows from the Glauber dynamics~\cite{glauber1963time}.
It can be shown that this process is Markovian and ergodic, and the system will eventually reach the thermal equilibrium.
To determine the speed of the convergence, we should notice the spectral gap of the transition matrix of the Markov chain, which is defined as

@@

Denote $S$ to be the length of the material along the ``Space'' axis, $D$
to be the length of the material along the ``Time'' axis. 
Let us define a simpler but equivalent energy model on a 2D lattice of size $S \times D$, whose hamiltonian
is given by
\begin{equation}\label{eq:toy-hamiltonian}
E(\vn) = \sum_{t=1}^{D-1}\sum_{i=1}^{S} (n_{i+1, t} \land \neg(n_{i, t} \land \neg n_{i-1, t})) \veebar n_{i,t+1},
\end{equation}

@@

\section{The implementation in Rydberg atoms array}\label{physical-model-rydberg-atoms-array}
Let us consider a Rydberg atoms array with a two dimensional embedding. The coordinate of the atom $v$ is denoted as $\mathbf{r}_v$.
The Rydberg Hamiltonian~\cite{Nguyen2023} is defined as
